Mg Depleted and K Enriched Stars from LAMOST Spectral Data

### 2018-01-20 article/ms.tex
Discussion refinements, and insertion of my thoughts of the implications of the gradient with [K/Fe] versus Fe

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}

\usepackage{color}
\usepackage{booktabs,chemformula}
\usepackage[export]{adjustbox}

@@

The data set of \LamostGiants\ \lamost\ giants was prepared by \citet{ho2017} for use with the supervised machine learning code \tc. The \lamost\ data was shifted to rest frame and pseudo-continuum-normalised before it was resampled to a common uniform wavelength grid between 3905\,\AA\ and 9000\,\AA. \tc\ was used to infer $\teff$, $\logg$, [Fe/H], and [$\alpha$/Fe] using 9,952 stars in common between \lamost\ and the higher spectral resolution \apogee\ survey through the use of a data-driven spectral model. For details regarding the preparatory work, model generation, and label transfer between the \apogee\ and \lamost\ surveys, the reader is directed to \citet{ho2017}.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{loggteffof113.png}
    \caption{Effective temperature, log surface gravity, metalicity}
    \label{tefflogg}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{KvsMgab.png}
    \caption{[Comparison of K vs Mg abundances with NGC 2419 and NGC 2808.}
    \label{KvsMg}
\end{figure}

@@

\section{Discussion}

\begin{itemize}
\item Association with globular clusters: basically all stars (will have to redo plot and search for the new tables) are not associated with globular clusters. in terms of those which are kinda close in GC, state which cluster(s) they could be associated with. We lack dynamics information, but potentially the next data release from GAIA will help with that. Check with apparent magnitudes to see if the stars are consistent in that regard.

\item Metalicity: the candidate stars are found at a wide range of metalicities, from \todo{-1.4<[Fe/H]<0.3}. for comparison, NGC 2808 and NGC 2419 have [Fe/H] ~-2.

\item Abundances: K abundances are lower and Mg abundances are higher than those in the extreme population of 2419. however, they \textit{are} still (perhaps) significantly K enhanced and Mg depleted, perhaps the first field stars identified as such. Only 113 out of \LamostGiants \ stars were found to be good candidates for enhanced K and reduced Mg. Likely not products of the same process as acted in NGC 2419, but possibly? Main implications: searched \LamostGiants \ stars, but didn't find any that we could say 'yeah that looks like it could have came from the same process that caused NGC 2419's signiature'.

@@

\item spread in Logg and Teff: the fact that we see stars across a range of Teff and Logg eliminates the idea that the candidates as all being so called 'red-clump' red giants or any other phenomena specific to certain RGB stages/ types.

@@

>>>>

\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{float}
\usepackage{gensymb}
\usepackage{color}
\usepackage{booktabs,chemformula}
\usepackage[export]{adjustbox}

@@

The data set of \LamostGiants\ \lamost\ giants was prepared by \citet{ho2017} for use with the supervised machine learning code \tc. The \lamost\ data was shifted to rest frame and pseudo-continuum-normalised before it was resampled to a common uniform wavelength grid between 3905\,\AA\ and 9000\,\AA. \tc\ was used to infer $\teff$, $\logg$, [Fe/H], and [$\alpha$/Fe] using 9,952 stars in common between \lamost\ and the higher spectral resolution \apogee\ survey through the use of a data-driven spectral model. Unless otherwise stated, these transfered labels are those used throughout this study. For details regarding the preparatory work, model generation, and label transfer between the \apogee\ and \lamost\ surveys, the reader is directed to \citet{ho2017}.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{loggteffof113.png}
    \caption{Effective temperature, log surface gravity, metalicity. \todo{The large, random spread in $\logg$, $\teff$ and [Fe/H] implies that the candidates are simply stars which have formed from poluted material, rather than any special type of star such as red-clump giants.}}
    \label{tefflogg}
\end{figure}

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{KvsMgab.png}
    \caption{[Comparison of K vs Mg abundances with NGC 2419 and NGC 2808. The dotted divisor line is present simply as a visual aide.}
    \label{KvsMg}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{KvsFe.png}
    \caption{[K/Fe] vs [Fe/H] for the LAMOST data}
    \label{KvsFe}
\end{figure}

\begin{figure}
	\includegraphics[width=\columnwidth]{NavsFe.png}
    \caption{[Na/Fe] vs [Fe/H] for the LAMOST data. \todo{NOTE: NONE OF THE Na ENHANCEMENTS ARE IN OUR MOST METAL POOR STARS. This is interesting because Na enhancements are associated with some globular cluster populations, but here we're seeing sodium enhancements in intermediate-high metalicity stars, with the non-Na population spanning the full range of metalicities.}}
    \label{NavsFe}
\end{figure}

@@

\section{Discussion}
\subsection{Association with globular clusters}

Of the 113 candidate sample identified, J074807.22+261514.2 appears closest in the sky to a globular cluster, and even then still at a significant angular separation of 2.54\degree . Although little is known about Ko 2 other than that it is an extremely faint halo cluster \citep{koposov2007}, it would be expected that a star from such a cluster would have very low metalicty, while J074807.22+261514.2 has $ \text{[Fe/H]} = -0.24 \pm 0.07$. Stars from Ko 2 would be unlikely to appear in a survey such as \lamost, which heavily favours detection of bright field stars.

None of the stars appear to be associated with NGC 2419, NGC 2808 or $\omega$ Centauri. Only 4 stars were also contained within \textit{GAIA}'s DR1, and when backwards time integrated using GAIA's data with the GALA package they exhibited orbits consistent with field stars, with none of the eccentricity and high velocity typical of a star disrupted from a globular cluster.

The spread of metalicity shown in Figure \ref{mhist} further supports the sample being made up of field stars, displaying a spread reflective of the original \LamostGiants \ sample of giants. If there was a significant influence in the sample from disrupted stars from another as yet unidentified globular cluster containing the Mg-K anti-correlation, a significant deviation in the metalicity histogram from the base \lamost \ sample of giants should be visible.

There is no evidence at this time supporting any stars from this sample having origins in globular clusters, while Figure \ref{galcord} clearly indicates galactic latitudes of the candidates are consistent with field stars in the disk of the galaxy. The spread in metalicity shown in Figure \ref{mhist}, with significant numbers of high metalicity stars, further supports the conclusion that the sample is comprised of field stars presumably formed at different times throughout the lifetime of the Milky Way.

\begin{itemize}
\item Association with globular clusters: basically all stars (will have to redo plot and search for the new tables) are not associated with globular clusters. in terms of those which are kinda close in GC, state which cluster(s) they could be associated with. We lack dynamics information, but potentially the next data release from GAIA will help with that. Check with apparent magnitudes to see if the stars are consistent in that regard.
\end{itemize}

\subsection{Abundances}

Figure \ref{KvsMg} shows [K/Fe]  for the 113 star sample plotted against $[\alpha$/Fe] overlaid with [K/Fe] from NGC 2419 and NGC 2808 plotted against [Mg/Fe] \citep{cohenkirby2012,mucciarelli2012,mucciarelli2015}. [K/Fe] vs [Mg/Fe] for the three stars from the sample observed using \textit{Magellan} are also plotted. $[\alpha$/Fe] is used as an upper limit for [Mg/Fe] for the 113 star sample as [Mg/Fe] proved impossible to obtain from the \lamost \ spectra. The use of [$\alpha$/Fe] as a proxy for [Mg/Fe] is supported by the [Mg/Fe] measurements from the high-resolution spectra obtained, these being similar to the [$\alpha$/Fe] abundances from \tc.

The candidates' [K/Fe] are generally lower than the extreme Mg-depleted population in NGC 2419, but higher than the Mg-normal population. It is perhaps noteworthy that both populations identified in NGC 2808 have significantly lower levels of [K/Fe] than these stars.
Levels of [$\alpha$/Fe] fall within the lower limits of the Mg-normal population of both clusters, although as this represents only an upper limit the exact level of Mg depletion for the candidates is unclear. The [Mg/Fe] abundances derived from \textit{Magellan}'s high resolution spectra imply that at most a small depletion relative to the Mg-normal population is present.

A similar relationship emerges when comparing with typical galactic figures for [K/Fe] and [Mg/Fe], taken from \cite{kobayashi2011}.
Only the [K/Fe] appears definitively anomalous, with galactic values for [K/Fe] around $0.3 \pm 0.2$ dex, while the \lamost \ sample of stars varies from around 0.8-1.3, and the \textit{Magellan} abundance for candidate star J091825.49+172114.5 having [K/Fe] $\approx$ 1.54.

The sample's [K/Fe] also exhibits an interesting anti-correlation with total metalicity, shown in Figure \ref{KvsFe}, which is absent from measurements in the Milky Way. This is consistent with a (somewhat) constant level of [K/H] between each star, causing [K/Fe]to fall with increasing metalicity.

In contrast, [$\alpha$/Fe] varies from around 0-0.3 dex in the \lamost \ sample, similar to galactic levels of [Mg/Fe], which vary with metalicity from approximately 0.4$\pm$0.2 for [Fe/H] $\sim$ -1.5 to 0.1 $\pm$0.2 for [Fe/H] $\sim$ -0.3 \citep{kobayashi2011}. No variation with metalicity is observed within the \lamost sample.

[Na/Fe] was also measured for the \lamost\ data, shown in Figure \ref{NavsFe}. \todo{See caption for relevant discussion point.}

The implications of the calculated abundances are intriguing. The fact that from such a large sample of field stars none were found that replicated the extreme abundance pattern of NGC 2419 further and perhaps definitively cements the clusters uniqueness.

However, the 113 star candidate pool of K enhanced and potentially Mg depleted stars identified represent a unique collection of stars; while not as K enhanced as the extreme population in NGC 2419, they exhibit K abundances well above typical galactic levels and a range of metalicities. It seems likely that whatever process is responsible for these anomalous field stars is similar to whatever caused the unusual stellar populations in NGC 2419 and NGC 2808.

\begin{itemize}
\item Abundances: K abundances are lower and Mg abundances are higher than those in the extreme population of 2419. however, they \textit{are} still (perhaps) significantly K enhanced and Mg depleted, perhaps the first field stars identified as such. Only 113 out of \LamostGiants \ stars were found to be good candidates for enhanced K and reduced Mg. Likely not products of the same process as acted in NGC 2419, but possibly? Main implications: searched \LamostGiants \ stars, but didn't find any that we could say 'yeah that looks like it could have came from the same process that caused NGC 2419's signiature'.

\subsection{A working hypothesis}
Significantly enhanced [K/Fe] is rare; significantly enhanced [K/Fe] without an accompanying depletion of [Mg/Fe] is unknown. Further, the candidate stars are spread across a wide range of metalicities.

This spread in metalicity is important; it strongly implies that the process responsible for this signature, rare as it is, is not necessarily tied to a certain epoch in the history of our galaxy. Rather, these stars appear to be continuously forming from successive generations of stars.

\todo{But... If real, the [K/Fe] trend with metalicity might actually imply the exact opposite. Perhaps a single event enriching a sufficiently large cloud of gas occurs. Stars are formed from the enriched gas, then die, releasing more metal to the gas, increasing iron etc, but leaving the hard-to-produce K abundance at around the same level relative to H, which would make up the majority of the gas. Then the successive stars formed from this material would have a lower value of [K/Fe], but a higher value of [Fe/H]. Problems this idea might have might include: how long a star forming gas ball can really last in the disk? Plus, if all these stars are forming in the same place, there sure is a lot of them and they don't seem to be associated with only one place, their all over the sky. This idea is actually fairly compatible with NGC 2419 though; one event (still utterly mysterious) providing material from which a generation of stars formed with enhanced K. Since it was just a globular cluster, and was lucky to have even 2 generations of stars, such a trend in metalicity would be non existent (seeing as there is no metalicity spread within the cluster anyway). In short, the metalicity gradient supports a single epoch of K formation at least, if not a single localised event, rather than the continuous production of K throughout the lifetime of the galaxy. The real question is whether or not it is real, something that we are unlikely to be able to answer.}

This immediately eliminates as a polluter candidate any theoretical hyper-massive Population III star that might be imagined. While such an object might be plausible in the context of NGC 2419, it cannot explain the enhanced K abundances observed in the disk at such a spread in metalicity.

As emphasized by \cite{prantzos2017}, a large part of the problem presented by the signatures of NGC 2419 and NGC 2808 is obtaining both a depletion Mg and an enhancement in K. Reducing Mg is possible through conversion of \ce{^{24}Mg} to \ce{^{23} Na} and through Mg-Al chains at temperatures around 75 MK, obtainable through hot-bottom burning in AGB stars while it takes temperatures upwards of 150 MK to produce K, temperatures potentially obtainable at the bottom of the envelope in SAGB stars \citep{iliadis2016}. This implies that the Mg depletion and K enrichment occur at separate sites.

One possible scenario that could produce a large K enhanced signature would be a binary in which one of the stars is a SAGB star. Then, if this SAGB star were to produce K, the star may accrete onto its smaller companion which would continue to evolve long after the SAGB star is gone. This companion would then, depending on its mass after accretion, either undergo hot-bottom burning or, more probably, would not. If the companion star had sufficient mass to undergo hot bottom burning, then Na should be enhanced from production from \ce{^24 Mg} and \ce{^22 Ne}.

\todo{Gain a much clearer picture about where Mg fits into all of this. NGC 2419 shows a very clear Mg division. However, there are only 2 generations of stars. I need to find out if its possible that such a clear Mg division would survive another generation; that is, when the Mg-depleted stars die, they release metals, including the Mg they formed in the core (assuming they were big enough). would this Mg released obliterate the signal after a generation? that would be nice and consistent with the data, and explain why we're not seeing a strong Mg depletion in our stars; that would only happen in the generation imediately following the enrichment of the gas with Mg-depleted material. Even better, because, unlike Mg which is easy to form, K is hard to form, the K signature in the material would be preserved through successively more metal rich generations of star forming material.}

This scenario allows for K abundances to appear at a range of metalicities, as this scenario could play out throughout the life of the Milky Way. It also  accounts for the significant minority population of the stars which show Na enhancements.

It does not, however, account for the uniqueness of the signal to NGC 2419 and NGC 2808 among globular clusters, nor does it satisfactorily explain why such a large proportion of stars in NGC 2419 exhibit the Mg-K anti-correlation.

\begin{itemize}

@@

@@

\subsection{My thoughts on an explanation (insatisfactory and incomplete regarding the identity of the poluter star(s)) consistent with the data}

Some unknown event or events that can only occur in very metal poor environments happens. This leaves gas which can form stars with depleted Mg and enhanced K.

The generation of stars that forms directly from this gas carries this signiature, and displays a significant reduction in Mg abundance and enhancement in K abundance similar to NGC 2419. Call this Generation 2.

When the big hot stars of Generation 2 die however, they release metals into the this gas, including Mg, which is easy to form, but NOT K, which is hard to form. The Mg depletion originally carried in this gas relative to the original levels of iron is now harder to detect, obscured by the influx of Mg from Generation 2, but the K enhancement is still relatively easy to detect, having only been reduced relative to iron by the total amount of iron released.

As this cycle continues, [K/Fe] would become lower and lower, approaching 'regular' galactic levels, while depletions in [Mg/Fe] would quickly very very faint.

This cycle allows for the Mg depletions to be preserved in globular clusters, where your lucky to get 2 generations of stars forming as the cluster lacks the gravity to hold on to sufficient gas for many generations of star. But it also allows the Mg signal to effectively be lost within the disc of the galaxy, where many generations of stars may form.

\begin{itemize}

\item spread in Logg and $\teff$: the fact that we see stars across a range of $\teff$ and Logg eliminates the idea that the candidates as all being so called 'red-clump' red giants or any other phenomena specific to certain RGB stages/ types.
